### 2021-06-17 memo.tex
anomaly from monopole

\def\bZ{\mathbb{Z}}
\def\fP{\mathfrak{P}}
\def\diag{\mathop{\mathrm{diag}}}

@@

\section{'t Hooft-Polyakov monopole argument}
For an $SU(2)$ gauge theory Higgssed by an adjoint ($\bm{3}$, spin-$1$, isovector) scalar,
the gauge group is broken down to $U(1)$, and correspondingly it accommodates topological solitons (monopoles):
\begin{equation*}
	\pi_2\left(
		\dfrac{SU(2)}{U(1)}
	\right) = \bZ.
\end{equation*}
In the presence of (additional) fermions, this monopole might acquire non-trivial charge under
spacetime-Lorentz or flavor symmetries, depending on the representation of the fermions under $SU(2)$ gauge symmetry:
\begin{equation*}
	\begin{tabular}{c||cccc}
		fermion gauge rep. & number of zero-modes & spin of zero-modes& spin of monopole\\
		\hline
		$\bm{2}$ & $1$ & $0$ & $0$\\
		$\bm{3}$ & $2$ & $\tfrac{1}{2}$\\
		$\bm{4}$ & $4$ & $0\oplus 1$ & $\tfrac{1}{2}$\\
	\end{tabular}
\end{equation*}
The numbers of zero-modes can be computed from the Callias index theorem\cite{Callias:1977kg}.

@@

Also, one should be able to generalize this whole argument to the case of fermions charged under
\begin{equation*}
	\dfrac{SO(2n_c)_{\text{gauge}} \times SU(2n_f)_{\text{flavor}}}{\bZ_2}
\end{equation*}
by breaking $SU(2n_f)_{\text{flavor}}$ to $SO(2n_f)_{\text{flavor}}$,
where we have monopoles characterized by
\begin{equation*}
	\pi_2\left(
		\dfrac{SU(2n_f)}{SO(2n_f)}
	\right)
	=
	\left\{
		\begin{array}{cl}
			\bZ_2 & (n_f \geq 2)\\
			\bZ & (n_f = 1)
		\end{array}
	\right..
\end{equation*}

>>>>

\def\bR{\mathbb{R}}
\def\bZ{\mathbb{Z}}
\def\fP{\mathfrak{P}}
\def\diag{\mathop{\mathrm{diag}}}

@@

\section{'t Hooft-Polyakov monopole argument}
\subsection{Basics}
For an $SU(2)$ gauge theory Higgssed by an adjoint ($\bm{3}$, spin-$1$, isovector) scalar,
the gauge group is broken down to $U(1)$, and correspondingly it accommodates topological solitons (monopoles):
\begin{equation*}
	\pi_2\left(
		\dfrac{SU(2)}{U(1)}
	\right) = \bZ.
\end{equation*}
In the presence of (additional) fermions, this monopole might acquire non-trivial charge under
spacetime-Lorentz or flavor symmetries, depending on the representation of the fermions under $SU(2)$ gauge symmetry:
\begin{equation*}
	\begin{tabular}{c||cccc}
		fermion gauge rep. & number of zero-modes & spin of zero-modes& spin of monopole\\
		\hline
		$\bm{2}$ & $1$ & $0$ & $0$\\
		$\bm{3}$ & $2$ & $\tfrac{1}{2}$\\
		$\bm{4}$ & $4$ & $0\oplus 1$ & $\tfrac{1}{2}$\\
	\end{tabular}
\end{equation*}
The numbers of zero-modes can be computed from the Callias index theorem\cite{Callias:1977kg}.

@@

\subsection{Generalization}
\subsubsection{Breaking pattern 1}
Also, one should be able to generalize this whole argument to the case of fermions charged under
\begin{equation*}
	\dfrac{SO(2n_c)_{\text{gauge}} \times SU(2n_f)_{\text{flavor}}}{\bZ_2}
\end{equation*}
by breaking $SU(2n_f)_{\text{flavor}}$ to $SO(2n_f)_{\text{flavor}}$,
where we have monopoles characterized by
\begin{equation*}
	\pi_2\left(
		\dfrac{SU(2n_f)}{SO(2n_f)}
	\right)
	=
	\left\{
		\begin{array}{cl}
			\bZ_2 & (n_f \geq 2)\\
			\bZ & (n_f = 1)
		\end{array}
	\right..
\end{equation*}

\subsubsection{Breaking pattern 2}
We can also consider $SO(2n_c) \times USp(2n_f)$ breaking to $SO(2n_c) \times U(n_f)$ 
where the $2n_f$ dimensional irrep of $USp(2n_f)$ becomes 
a fundamental plus an antifundamental of $U(n_f)$.

The induced effective interaction is \begin{equation}
\frac12 w_2(SO(2n_c)) c_1(U(n_f)).
\end{equation} 
(In the following we consider cochains valued in $\bR/\bZ$.)

We now take the $\bZ_2$ quotient.
Note that $\pi_1(U(n_f)/\bZ_2)$ is $\bZ\times \bZ_2$  or $\bZ$ depending on whether $n_f$ is even or odd.
Furthermore, when $n_f$ is even, $c_1(U(n_f)) = c_1(U(n_f)/\bZ_2)$ and $v_2(U(n_f)/\bZ_2)=v_2(USp(2n_f))$/
When $n_f$ is odd, $c_1(U(n_f)/\bZ_2)=2 c_1(U(n_f))$ when the latter is well-defined and $v_2(U(n_f)/\bZ_2)$ is the mod-2 reduction of $c_1(U(n_f)/\bZ_2)$.

We now compute the anomaly cochains in the four cases separately:
\paragraph{$(n_c,n_f)$=(even,even):}
$w_2(SO(2n_c))$ and $c_1(U(n_f))$ can be generalized without problem,
and therefore \begin{equation}
\delta(\frac12 w_2(SO(2n_c)) c_1(U(n_f))) = 0.
\end{equation}

\paragraph{$(n_c,n_f)$=(odd,even):}
$w_2(SO(2n_c))$ needs to be upgraded to a $\bZ_4$-valued cochain $a_2(SO(2n_c)/\bZ_2)$.
The original interaction is then \begin{equation}
\frac14 a_2(SO(2n_c)/\bZ_2) c_1(U(n_f))
\end{equation}
which is closed without problem.

\paragraph{$(n_c,n_f)$=(even,odd):}
Here we need to replace $c_1(U(n_f))$ by $c_1(U(n_f)/\bZ_2)/2$.
The effective interaction is then \begin{equation}
\frac14 w_2(SO(2n_c)/\bZ_2) c_1(U(n_f)/\bZ_2) 
\end{equation} whose $\delta$ is \begin{align}
\delta(\frac14 w_2(SO(2n_c)/\bZ_2) c_1(U(n_f)/\bZ_2) )
&= \frac12 (\frac12\delta w_2(SO(2n_c)) c_1(U(n_f)/\bZ_2) ) \\
&= \frac12 (\beta w_2(SO(2n_c)))c_1(U(n_f)/\bZ_2) 
\end{align}
which is the pull-back of the anomaly cochain
\begin{equation}
 \frac12 (\beta w_2(SO(2n_c)))v_2(USp(2n_f)/\bZ_2). 
\end{equation}

\paragraph{$(n_c,n_f)$=(odd,odd):}
Now we make the replacement on both sides and the effective interaction is \begin{equation}
\frac18 a_2(SO(2n_c)/\bZ_2) c_1(U(n_f)/\bZ_2) 
\end{equation}
whose $\delta$ is \begin{align}
\delta(\frac18 a_2(SO(2n_c)/\bZ_2) c_1(U(n_f)/\bZ_2) )
&= \frac12 (\frac14\delta a_2(SO(2n_c)/\bZ_2) c_1(U(n_f)/\bZ_2) ) \\
&= \frac12 (\beta_2 a_2(SO(2n_c)/\bZ_2))c_1(U(n_f)/\bZ_2) 
\end{align}
which is the pull-back of \begin{equation}
 \frac12 (\beta_2 a_2(SO(2n_c)))v_2(USp(2n_f)),
\end{equation}
which is more or less what we need, 
since $a_2(SO(2n_c)=v_2(USp(2n_f)/\bZ_2)$ up to exact terms.
